Locality and Computational Reliability of Linear Response Calculations for Molecular Systems

### 2018-11-19 AnalyticStructure_main.tex
removed changes and start review of the text

% Useful packages
\usepackage{graphicx}
\usepackage{subfig}
\usepackage{amsmath,amsfonts,amssymb,amsthm}
\usepackage{mathtools,braket}
\usepackage{tikz}
\usepackage{xifthen}
%\pgfplotsset{compat=1.15}
%\usepackage[markup=underline]{changes}
\usepackage[final]{changes}
\definechangesauthor{dal}
\definechangesauthor{gen}

@@

\section{Introduction}
\deleted[id=gen]{%
Since the days of their foundations, disciplines like Computational Physics or Quantum Chemistry always had to deal with the problem of the
computational reliability of the results. More specifically, a computational result may be considered reliable for various reasons.
For example, it may be in agreement with some previous observation or measurement that the employed computational model is susceptible to reproduce.
The model itself may in this way be classified as ``accurate''; it can then be employed to extract, \textit{in silico},
other quantities that might be then used to \emph{predict} or \emph{interpret} experiments.
Another concept of reliability is the concept of  ``precision'', which is more focused to the actual results
that a given computational model provides for a specific set of quantities, regardless of their potential agreement with
experimental data. The aim of precise computational approach is to reduce the \emph{computational} uncertainties
of quantities extracted out of a well-defined model, and provide reference results, to which other computer codes
employing the same model can be compared.
%
A recent, remarkable example of precision-driven initiative is provided by the DeltaCodes project~\cite{deltaTest2016},
where a grassroot community of code-owners struggled to extract well-defined computational results in the context of
Density Functional Theory (DFT) calculations. Ground-State quantum mechanical quantities as lattice constants and bulk modulus of
elementary crystals extracted from different codes were compared with each other, with the objective of reducing the uncertainties
on such data, and in some sense, to define a \emph{calibration} procedure for a Solid-State Physics DFT code.
Any other code able to extract the same quantities with the same approximations may be compared with the
results of the community in order to assess its computational reliability.
}

@@

In this paper, we focus our attention to Linear-Response treatments of \emph{molecular} systems,
in particular referring to LR-TDDFT calculation. We will investigate the impact that
\emph{open} (i.e. Isolated) Boundary Conditions (BC) that have to be imposed to the system will have on the computational evaluation of LR quantities.
\deleted[id=gen]{In order to do this}We perform analytic considerations on the equations beyond LR treatment and discuss the interplay between reproducible LR quantities and the \emph{locality} of the related objects. Such considerations will enable us to distinguish quantities that \textit{a priori} can be compared among different treatments - and therefore tested for reproducibility - from others that would \emph{explicitly} depend of the employed computational treatment.

It is indeed relatively easy to have converged indicators for numerical Ground State (GS) quantities: on one side, the variational theorem
guarantees that the lower the GS energy, the more precise the numerical result;
on the other side, GS quantities of molecular system  can be expressed
only in terms of localized (i.e. bound) states. Such considerations make relatively easy the increasing  of the basis set completeness in view of the extraction of reference results.

We will see in what follows that the situation is not as simple in the context of LR.
Here we will show that the response density -- even for molecular systems -- looses its localized behaviour for high frequencies,
independently on the nature of the system and of the perturbation.
Furthermore, we will demonstrate that to increase basis set completeness, delocalized oscillatory degrees of freedom cannot be
excluded \textit{a priori}, even in the localized regime.

We will also see that the reliabilty of the convergence criteria that can be established for LR -- where no variational theorem exists -- depends explictily on the target frequency range and on the features of the numberical basis set employed.

@@

\replaced[id=dal]{Let us start our discussion by reviewing some well-known concepts of quantum mechanic under a perspective that
will turn out to be very useful in the forthcoming considerations about linear-response quantities. In particular we revisit the impact
that the Isolated BC have on the solutions of the Schr\"odinger Equation.}{Let us start our discussion by revisiting the
impact that the Isolated BC have on the solutions of the time-independent Schr\"odinger Equation. We here present well-known concepts of
quantum mechanics, under a perspective that will turn out to be very useful in the forthcoming considerations about linear-response quantities.}

@@

We can therefore introduce an explicit representation of the
response density parametrized as
\be\lb{rhoPrimeFluctuationStateDef1}
\dm'_\Phi(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p^\Phi(-\omega)} + \ketbra{f_p^\Phi(\omega)}{\psi_p}\right) \;.
\ee
Here we have introduced the set of $\omega$-dependent \emph{fluctuation states} (FS)
$\ket{f_p^\Phi(\omega)} = \dm'_\Phi(\omega) \ket{\psi_p}$, that have values in the unoccupied subspace of $\hnot$.
\added[id=gen]{
However, non-hermiticity of $\dm'_\Phi(\omega)$ implies that the
fluctuation states are complex quantities.}
%  Causality implies that the real and imaginary part of FS are related each other by the Kramers-Kronig relations}
% \be\lb{KKrel}
% \mathrm{Re}\left( \ket{f_p^\Phi(\omega)} \right) =
% \mathcal P \int \dd \omega'
% \frac{\mathrm{Im}\left(\ket{f_p^\Phi(\omega')}\right)}{\omega -\omega'}.
% \ee
% \added[id=gen]{We will come to these equations in the forthcoming sections.}
%}

@@

\deleted[id=dal]{Modified the name of the subsection}

@@

%\subsection{Strong and weak convergence at finite $\omega$}

The analysis described above has shown that the building blocks of the response density $\{\ket{f_p(\omega)}\}$
are susceptible to exhibit bound-state long-range behavior \emph{only} when $\omega < |\eps_p|$,
while they behave as unbound states for values of $\omega$ above the threshold levels.
This fact has interesting implications in the evaluation of LR quantities. Indeed, writing the response
density in terms of FS implies that \deleted[id=dal]{(equation for the response density modified)}
% \be\lb{LinearResponseFunctDef2}
% \braket{\delta\op O}_\Phi(\omega) = 2 \mathrm{Re} \left(\sum_p \bra{\psi_p}\op O\ket{f_p^\Phi(\omega)}\right) \;.
% \ee
\begin{multline}\lb{LinearResponseFunctDef2}
\braket{\delta\op O}(\omega) = \sum_p \left( \bra{f_p^\Phi(-\omega)}\op O\ket{\psi_p}+ \bra{\psi_p}\op O\ket{f_p^\Phi(\omega)}\right) \\
=2 \sum_p\bra{\psi_p}\op O\ket{f_p^\Phi(\omega)}  \;.
\end{multline}
In the second equality we have assumed that $\op O$ is a real symmetric operator and chosen a real set of occupied molecular states.
Let us suppose that also the observable of interest $\op O$ \replaced[id=gen]{may be restricted to localized domains in the same way as the potentials $ \op V$ and $\op \Phi$.}{exhibits the locality properties introduced for the potentials $ \op V$, $\op \Phi$ and $\op V'$.}
Once again, it would be enough to consider a local operator for this condition to be met.

Two
distinct regimes can be identified, on the basis of value of $\omega$.
The first one, which we will refer to as the \emph{below threshold} regime, is realized when
$\omega$ is lower than the ionization potential $|\eps_h|$.
In this case all the FS are below their threshold level and equation \eqref{LinearResponseFunctDef2}
is expressed in terms of genuinely localized
quantities. At these frequencies, the linear response functional $\braket{\delta\op O}_\Phi(\omega)$
may thus be evaluated with computational setups that
are similar to the ones usually employed for GS calculations.
The convergence in this regime is ``strong'', in the sense that the basis functions employed enables us to express
with arbitrary precision \emph{both} the states $\ket{\op O \psi_p}$ and $\ket{f_p^\Phi(\omega)}$.
This is of course true even when the above states are only implicitly expressed by the employed LR treatment,
 as e.g. in the case of Krylov spaces generated from the states $\ket{\op\Phi(\omega) \psi_p}$ \cite{baroni2006,baroni2008,linlinKPM}.
 \added[id=gen]{In this regime, a computational treatment based on localized basis set is susceptible to provide a precise answer, assuming a reasonable level of completeness.}

On the contrary, the computation of the linear response functional can be much more demanding in the \emph{above threshold regime}, realized for $\omega>|\eps_h|$, in which fluctuation
states start behaving as unbound wave functions.
As by hypothesis $\ket{\op O \psi_p}$ is bound-state like, \emph{only} the scalar products of Eq.~\eqref{LinearResponseFunctDef2} can be evaluated in a localized domain.
In this case a computational description of fluctuation states is only meaningful in ``weak'' sense,
where the bound state character of $\{\bra{\psi_p}\op O\}$ behaves as a regulator.
This is naturally achieved in the above mentioned Krylov space treatments, which
explicitly deal with the expression of the scalar products.
\added[id=gen]{However, in this regime, imposing \emph{by design} a localized
behaviour of the FS might reveal very dangerous in view of convergence of the results:
the computational basis set have to be able to express in the support domain of $\{\bra{\psi_p}\op O\}$ a asymptotically delocalized FS. The degrees of completeness
of the computational basis should therefore be increased by imposing oscillatory degrees of freedom needed to represent the FS at the border of the simulation domain: it is important that the underlying basis set is able to express delocalised states.}

@@

As an example in support of these arguments we compare the \replaced[id=dal]{dynamical polarizability $\alpha(\omega)$}{absorption spectrum} of a $CO$ molecule computed in three distinct real-space computational setups, which differs for the dimension of the computational domain.
Results reported in \replaced[id=dal]{the two panels of figure \ref{co_spectrum} show the real and imaginary part of $\alpha(\omega)$, respectively. In all the cases a typical behavior emerges}{evidences a typical behavior}: the curves are almost coincident in the below threshold regime,
whereas the effect of the computational setup are clearly visible for values of  $\omega$ that exceed the threshold value.
For these high-frequency regions, convergence might be reached for domains of much larger sizes than what presented here~\cite{baroni2008},
which are -- as already noticed -- more than sufficient below threshold.

@@

\added[id=dal]{It is possible to reformulate this type of analysis by directly looking at LR quantities that are defined independently from the perturbing operators and target observables. This can be achieved by studying the
\emph{susceptibility functional}, defined implicitly trough the operator equation}
\deleted[id=dal]{
In view of reproducibility, the scalar product of Eq.~\eqref{LinearResponseFunctDef1}, defining the observable fluctuation, may be compared among various numerical treatments.
However, even though directly related to experimental results, such quantity is not by itself a property of the molecule, as it is a function of $\omega$ that explicitly depends
on the perturbation operator $\op \Phi$, not to mention the target observable $\op O$. Alternatively, we may consider only the fluctuation state, but it is still a $\op \Phi$-dependent wavefunction,
with a potential local behavior only for $\omega$ below the ionization threshold. Hence, comparing fluctuation states among diverse numerical treatments would be a
challenging task.}
\deleted[id=dal]{It is therefore much more interesting to concentrate on LR quantities that are defined independently from the perturbing operators and target observables.
This can be achieved by studying the
\emph{susceptibility functional}, defined implicitly trough the operator equation}
\be\lb{SusceptibilityDef1}
\dm'(\omega) = \left(\omega -\mathcal L \right)^{-1} \commutator{\op \Phi(\omega)}{\dmnot} = \sop{{\cal \chi}}(\omega) \op\Phi(\omega) \;.
\ee
The susceptibility (super)operator $\cal \chi(\omega)$ is therefore closely related to the resolvent of the Liouvillian of Eq.~\eqref{LiouvillianRhopomegaDef1}, which is completely defined
from the molecule Hamiltonian $\hnot[\dm]$.
The ``excitation modes''  of the molecule are defined through the \emph{excitation operators}, satisfying
\be\lb{ExcitationOperatorsDef1}
\Liouv \op E_a = \Omega_a \op E_a \qq \op{\tilde E}_a \Liouv = \Omega_a \op{\tilde E}_a \;,
\ee
together with the operator orthonormalization condition %$\trace{\op{\tilde E}\op E'} = \delta(E-E')$.
\be\lb{orthoExcitatioOpDef1}
\trace{\op{\tilde E}_a\op E_b} = \delta_{ab} \;.
\ee
Excitations defined in that way may be considered as a basis of the operators that satisfy the transverse condition as in
Eq.~\eqref{RhopTransverseDef1}, and enable us to express the susceptibility as a spectral decomposition:
\be\lb{RhopExcitationDef1}
\sop{{\cal \chi}}(\omega) \cdot   = \sum_{\{a\}} \op E_{a} \,
\frac{\trace{\commutator{\op{\tilde E}_a}{\dmnot}\cdot}}{\omega-\Omega_a} \;.
\ee
In this way Eq.~\eqref{LinearResponseFunctDef2} may be also written as:
\begin{equation}\lb{deltaoexc}
  \braket{\delta \op O} =  \sum_{\{a\}} \frac{\trace{\op O \sop B_a \op \Phi(\omega)}}{\omega -\Omega_a}
\end{equation}
where the action of the spectral operator $\sop B_a$ reads
\begin{equation}
  \trace{\op O \sop B_a \op \Phi(\omega)} = \trace{\op O \op E_a}
  \trace{\op{\tilde E}_a \commutator{\dmnot}{\op \Phi(\omega)}}
\end{equation}

\deleted[id=dal]{Modified name of the subsection}

@@

\added[id=dal]{The properties of the Liouvillian superoperator $\Liouv$ (sketched in appendix \ref{LiouvillianAction}) imply that both the left and right operator-valued
eigenstates \eqref{ExcitationOperatorsDef1} satisfy the transverse condition \eqref{RhopTransverseDef1} and can be parametrized as}
\begin{align}\lb{ExcitationOperatorsDef2}
\op E_a &= \sum_p \left( \ket{\phi^a_p}\bra{\psi_p} + \ket{\psi_p} \bra{\chi^a_p}\right)\;, \nn \\
\op{\tilde E}_a = \commutator{\dmnot}{\op E^t_a} &= \sum_p \left( \ket{\psi_p}\bra{\phi^a_p} - \ket{\chi^a_p} \bra{\psi_p}\right)\;.
\end{align}
\added[id=dal]{Each excitation ``mode'' of the system with associated energy $\Omega_a$ is thus described by a set of states $\{\phi^a_p,\chi^a_p\}_{p \in a}$, defined in the unoccupied subspace. These objects represent, respectively, the state in which $\ket{\psi_p}$
is excited -- or from whom it decays -- when the system is subject to the \emph{monochromatic} perturbation $\op \Phi_a \equiv \commutator{\dmnot}{\op E_a}$.}
% Each excitation ``mode'' of the system $\op E_a$ with energy $\Omega_a$, can be associated to a set of states $\{\phi^a_p,\chi^a_p\}_{p \in a}$,
% entirely defined in the unoccupied subspace.
% In this way we can parametrize, without loss of generality, the excitation operators of the system as:
% \begin{align}\lb{ExcitationOperatorsDef2}
% \op E_a &= \sum_p \left( \ket{\phi^a_p}\bra{\psi_p} + \ket{\psi_p} \bra{\chi^a_p}\right)\;, \nn \\
% \op{\tilde E}_a = \commutator{\dmnot}{\op E^t_a} &= \sum_p \left( \ket{\psi_p}\bra{\phi^a_p} - \ket{\chi^a_p} \bra{\psi_p}\right)\;.
% \end{align}
% For each choice of $\{\phi^a_p,\chi^a_p\}_{p \in a}$, these objects represent, respectively, the state in which $\ket{\psi_p}$
% is excited -- or from whom it decays -- when the system is subject to the \emph{monochromatic} perturbation $\op \Phi_a \equiv \commutator{\dmnot}{\op E_a}$.
We denote with the shorthand $p \in a$ the fact that the sum of Eq.~\eqref{ExcitationOperatorsDef2}
is in general restricted only to a $a$-dependent subset of the occupied orbitals, depending on the particular symmetry
of the excitation.
Excited states  satisfy the normalization condition
\be\lb{ExcitedStateOrthNormDef1}
\sum_p \left(\brket{\phi_p^a}{\phi_p^b} - \brket{\chi_p^b}{\chi_p^a}\right) = \delta_{ab} \;.
\ee
By applying \eqref{ExcitationOperatorsDef1} to the above parametrization
of the excitations, we obtain the following Sternheimer-like equations for the excited states:
\begin{align}\lb{ExcitationOperatorsDef3}
&\left[\Omega_a-(\hnot - \eps_p)\right] \ket{\phi_p^a} = \op Q_0 \op V'[\op E_a] \ket{\psi_p} \;, \nn\\
&\bra{\chi_p^a}\left[-\Omega_a-(\hnot - \eps_p)\right] = \bra{\psi_p} \op V'[\op E_a] \op Q_0  \;.
\end{align}
By writing these equations in a basis set of virtual states we obtain the well-known Casida's eigenvalue equation\cite{CasidaBook}
associated to the excitation energy $\Omega_a$ (see appendix \ref{casida}).
The spectrum is symmetric w.r.t the inversion of the eigenvalues
$\Omega_a \rightarrow -\Omega_a$ and \replaced[id=dal]{given a specific excitation $\{\phi^a_p,\chi^a_p\}$ the associated solution with opposite  energy is described by the transposed pair $\{\chi^a_p,\phi^a_p\}$.}{the associated excitation with negative energy are described by the transposed pair $\{\chi^a_p,\phi^a_p\}$.}
We concentrate our analysis on the sector of positive energies, i.e. $\Omega_a > 0$.

@@

The susceptibility functional has therefore a peculiar analytic structure, that
is made of multiple threshold energies, one for each of the ionization potentials
associated to the occupied states.
If an excitation $\op E_a$ has an energy $\Omega_a < |\eps_a|$ with $\eps_a = \mathrm{max}\left(\eps_p\right)_{p\in a}$,
it may belong to a discrete spectrum; otherwise it \emph{has} to belong to a continuum of states.
The typical structure of the excitation landscape is depicted in figure \ref{ExcitationLandscape}, once again assuming \replaced[id=dal]{that the perturbed potential $\op V'[\op E]$ gives rise to a bound state when applied to $\ket{\psi_p}$.}{a reasonable behavior of the perturbed potential $\op V'$.}
Below the first ionization energy $|\eps_h|$ only localized excitations are present. Conversely, in the intermediate energy interval that extends from $|\eps_h|$ to the
absolute value of the deepest occupied orbital, other discrete excitations may exist, and are embedded in a continuum of delocalized excitations.
For energies higher than $|\eps_1|$ \emph{no localized} excitations are anymore possible,
regardless of the particular behavior of $\op V'$: the excitation landscape in this region \replaced[id=dal]{contains a continuum set of delocalized states.}{only contains continuum states.}

\vspace{1cm}
\added[id=dal]{A proposal to reorganize the material of the following two subsection B e C in a different way}

\section{Physical relevance of the excitation spectrum. A comparison among localized and continuum sectors}
\deleted[id=dal]{
Comments on the features and the physical relevance of the two sector of the excitation spectrum}

%\subsection{Localized sector}

%\begin{itemize}
% \deleted[id=dal]{\item It is composed by a genuine set of discrete elements.}
% \deleted[id=dal]{\item The values of the excitation energies do not depend on the computational setup (observable features) and so can be compared among different approaches.}
% \deleted[id=dal]{\item The knowledge of this sector of the excitation spectrum allows us to express only the imaginary part of (proper equation that expresses the response functional written using Sokowsky-Plemely
% on the susceptibility functional).}
% \deleted[id=dal]{\item A computational setup in which delocalized degrees of freedom are excluded by design can in principle describe to this sector in a exact way.}
%\end{itemize}

\subsection{Continuum sector}

\begin{itemize}
   \item Make some statements on the features that a computational setup must have to guarantees that this sector provides converged results. Difference between systematic and localized basis setup. Can we comment about the needs to add ghost atoms for the evaluation of statical polarizability with gaussian basis?
 % \deleted[id=dal]{\item Not true poles of the susceptibility (but only an effective tool to compute the spectral function).}
 % \deleted[id=dal]{\item Deeply affected by the computational setup used to represent them. In particular, the presence of a finite computational domain implies a fictitious discretization of the excitation levels and the associated spectrum exhibits an explicit dependence from the size of the domain.}
 % \deleted[id=dal]{\item This sector of the excitation spectrum is needed to compute imaginary parts above IP and real part for \emph{all} the values of $\omega$. Add examples on the statical polarizability. Can we comment}
\end{itemize}

\vspace{1cm}

@@

The analysis of the above section reveals that the excitation spectrum is composed by two distinct sectors.
The first one is realized by the (finite) set of localized excitations with discrete eigenvalues.
Borrowing the terminology introduced in section \ref{SEopenSystem} to the present case, we can affirm that such localized excitations behave as
\emph{observable} quantities and the associated energies possess ideal features of reproducibility:
if the numerical basis employed is complete enough to
represent the states $\{\phi_p^a,\chi_p^a\}$, the energy $\Omega_a$
converges to its reference value.
\added[id=gen]{A computational setup in which delocalized degrees of freedom are excluded by design can, in principle, describe excitations of this sector in a exact way.}

>>>>

% Useful packages
\usepackage{graphicx}
\usepackage{subfig}
\usepackage{amsmath,amsfonts,amssymb,amsthm}
\usepackage{mathtools,braket}
\usepackage{tikz}
\usepackage{xifthen}
%\pgfplotsset{compat=1.15}

%\usepackage[markup=underline]{changes}
%\usepackage[final]{changes}
%\definechangesauthor{dal}
%\definechangesauthor{gen}

@@

\section{Introduction}

@@

In this paper, we focus our attention to Linear-Response treatments of \emph{molecular} systems,
in particular referring to LR-TDDFT calculation. We will investigate the impact that \emph{open} (i.e. Isolated) Boundary Conditions (BC), that are imposed to the system, will have on the computational evaluation of LR quantities.
We perform analytic considerations on the equations beyond LR treatment and discuss the interplay between reproducible LR quantities and the \emph{locality} of the related objects. Such considerations will enable us to distinguish quantities that \textit{a priori} can be compared among different treatments - and therefore tested for reproducibility - from others that would \emph{explicitly} depend of the employed computational treatment.

It is indeed relatively easy to have converged indicators for numerical Ground State (GS) quantities: on one side, the variational theorem
guarantees that the lower the GS energy, the more precise the numerical result; on the other side, GS quantities of molecular system  can be expressed
only in terms of localized (i.e. bound) states. These considerations make relatively easy to increase the basis set completeness in view of the extraction of reference results.

We will see in what follows that the situation is not as simple in the context of LR. Indeed we will show that the response density -- even for molecular systems -- looses its localized behaviour for high frequencies, independently on the nature of the specific system and perturbation. Furthermore, we will demonstrate that to increase the basis set completeness, delocalized oscillatory degrees of freedom cannot be excluded \textit{a priori}, even in the localized regime.

We will also see that the reliabilty of the convergence criteria that can be established for LR -- where no variational theorem exists -- depends explicitly on the target frequency range and on the features of the numerical basis set.

@@

Let us start our discussion by reviewing some well-known concepts of quantum mechanic under a perspective that will turn out to be very useful 
in the forthcoming considerations about linear-response quantities. In particular we revisit the impact that the Isolated BC have on the solutions 
of the Schr\"odinger Equation.

@@

We can therefore introduce an explicit representation of the
response density parametrized as
\be\lb{rhoPrimeFluctuationStateDef1}
\dm'_\Phi(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p^\Phi(-\omega)} + \ketbra{f_p^\Phi(\omega)}{\psi_p}\right) \;.
\ee
Here we have introduced the set of $\omega$-dependent \emph{fluctuation states} (FS)
$\ket{f_p^\Phi(\omega)} = \dm'_\Phi(\omega) \ket{\psi_p}$, that have values in the unoccupied subspace of $\hnot$.
However, non-hermiticity of $\dm'_\Phi(\omega)$ implies that the
fluctuation states are complex quantities.

@@

@@

The analysis described above has shown that the building blocks of the response density $\{\ket{f_p(\omega)}\}$
are susceptible to exhibit bound-state long-range behavior \emph{only} when $\omega < |\eps_p|$,
while they behave as unbound states for values of $\omega$ above these threshold levels.
This fact has interesting implications in the evaluation of LR quantities. Indeed, writing the response
density in terms of FS implies that 
\be\lb{LinearResponseFunctDef2}
\braket{\delta\op O}_\Phi(\omega) = 2 \sum_p\bra{\psi_p}\op O\ket{f_p^\Phi(\omega)}  \;.
\ee
% \begin{multline}\lb{LinearResponseFunctDef2}
% \braket{\delta\op O}_\Phi(\omega) = \sum_p \left( \bra{f_p^\Phi(-\omega)}\op O\ket{\psi_p}+ \bra{\psi_p}\op O\ket{f_p^\Phi(\omega)}\right) \\
% =2 \sum_p\bra{\psi_p}\op O\ket{f_p^\Phi(\omega)}  \;.
% \end{multline}
In the second equality we have assumed that $\op O$ is a real symmetric operator and chosen a real set of occupied molecular states.
Let us suppose that also the observable of interest $\op O$ can be restricted to localized domains in the same way as the potentials $ \op V$ and $\op \Phi$.
Once again, it would be enough to consider a local operator for this condition to be met.

Two distinct regimes can be identified, on the basis of value of $\omega$.
The first one, which we will refer to as the \emph{below threshold} regime, is realized when
$\omega$ is lower than the ionization potential $|\eps_h|$.
In this case all the FS are below their threshold level and equation \eqref{LinearResponseFunctDef2}
is expressed in terms of genuinely localized
quantities. At these frequencies, the linear response functional $\braket{\delta\op O}_\Phi(\omega)$
may thus be evaluated with computational setups that
are similar to the ones usually employed for GS calculations.
The convergence in this regime is ``strong'', in the sense that the basis functions employed enables us to express
with arbitrary precision \emph{both} the states $\ket{\op O \psi_p}$ and $\ket{f_p^\Phi(\omega)}$.
This is of course true even when the above states are only implicitly expressed by the employed LR treatment,
 as e.g. in the case of Krylov spaces generated from the states $\ket{\op\Phi(\omega) \psi_p}$ \cite{baroni2006,baroni2008,linlinKPM}.
 In this regime, a computational treatment based on localized basis set is susceptible to provide a precise answer, assuming a reasonable level of completeness.

On the contrary, the computation of the linear response functional can be much more demanding in the \emph{above threshold regime}, realized for $\omega>|\eps_h|$, in which fluctuation
states start behaving as unbound wave functions.
As by hypothesis $\ket{\op O \psi_p}$ is bound-state like, \emph{only} the scalar products of Eq.~\eqref{LinearResponseFunctDef2} can be evaluated in a localized domain.
In this case a computational description of fluctuation states is only meaningful in ``weak'' sense, where the bound state character of $\{\bra{\psi_p}\op O\}$ behaves as a regulator. This is naturally achieved in the above mentioned Krylov space treatments, which explicitly deal with the expression of the scalar products. However, in this regime, imposing \emph{by design} a localized behaviour of the FS might reveal very dangerous in view of convergence of the results:
the computational basis set has to be able to express in the support domain of $\{\bra{\psi_p}\op O\}$ a asymptotically delocalized FS. The degrees of completeness of the computational basis should therefore be increased by imposing oscillatory degrees of freedom needed to represent the FS at the border of the simulation domain: it is important that the underlying basis set is able to express delocalized states.

@@

As an example in support of these arguments we compare the {dynamical polarizability $\alpha(\omega)$} of a $CO$ molecule computed in three distinct real-space computational setups, which differs for the dimension of the computational domain.
Results reported in the two panels of figure \ref{co_spectrum} show the real and imaginary part of $\alpha(\omega)$, respectively. In all the cases a typical behavior emerges: the curves are almost coincident in the below threshold regime,
whereas the effect of the computational setup are clearly visible for values of  $\omega$ that exceed the threshold value.
For these high-frequency regions, convergence might be reached for domains of much larger sizes than what presented here~\cite{baroni2008},
which are -- as already noticed -- more than sufficient below threshold.

@@

It is possible to reformulate this type of analysis by directly looking at LR quantities that are defined independently from the perturbing operators and target observable. This can be achieved by studying the \emph{susceptibility functional}, defined implicitly trough the operator equation
\be\lb{ExcitationOperatorsDef1}
\Liouv \op E_a = \Omega_a \op E_a \qq \op{\tilde E}_a \Liouv = \Omega_a \op{\tilde E}_a \;,
\ee
together with the operator orthonormalization condition %$\trace{\op{\tilde E}\op E'} = \delta(E-E')$.
\be\lb{orthoExcitatioOpDef1}
\trace{\op{\tilde E}_a\op E_b} = \delta_{ab} \;.
\ee
Excitations defined in that way may be considered as a basis of the operators that satisfy the transverse condition as in
Eq.~\eqref{RhopTransverseDef1}, and enable us to express the susceptibility as a spectral decomposition:
\be\lb{RhopExcitationDef1}
\sop{{\cal \chi}}(\omega) \cdot   = \sum_{\{a\}} \op E_{a} \,
\frac{\trace{\commutator{\op{\tilde E}_a}{\dmnot}\cdot}}{\omega-\Omega_a} \;.
\ee
In this way Eq.~\eqref{LinearResponseFunctDef2} may be also written as:
\be\lb{deltaoexc}
  \braket{\delta \op O} =  \sum_{\{a\}} \frac{\trace{\op O \sop B_a \op \Phi(\omega)}}{\omega -\Omega_a}\;,
\ee
where the action of the spectral operator $\sop B_a$ reads
\be
  \trace{\op O \sop B_a \op \Phi(\omega)} = \trace{\op O \op E_a}
  \trace{\op{\tilde E}_a \commutator{\dmnot}{\op \Phi(\omega)}} \;. \nn
\ee

@@

The properties of the Liouvillian superoperator $\Liouv$ (sketched in appendix \ref{LiouvillianAction}) imply that both the left and right operator-valued
eigenstates \eqref{ExcitationOperatorsDef1} satisfy the transverse condition \eqref{RhopTransverseDef1} and can be parametrized as
\begin{align}\lb{ExcitationOperatorsDef2}
\op E_a &= \sum_p \left( \ket{\phi^a_p}\bra{\psi_p} + \ket{\psi_p} \bra{\chi^a_p}\right)\;, \nn \\
\op{\tilde E}_a = \commutator{\dmnot}{\op E^t_a} &= \sum_p \left( \ket{\psi_p}\bra{\phi^a_p} - \ket{\chi^a_p} \bra{\psi_p}\right)\;.
\end{align}
Each excitation ``mode'' of the system, with associated energy $\Omega_a$, is thus described by a set of states $\{\phi^a_p,\chi^a_p\}_{p \in a}$, defined in the unoccupied subspace. These objects represent, respectively, the state in which $\ket{\psi_p}$ is excited -- or from whom it decays -- when the system is subject to the \emph{monochromatic} perturbation $\op \Phi_a \equiv \commutator{\dmnot}{\op E_a}$. We denote with the shorthand $p \in a$ the fact that the sum of Eq.~\eqref{ExcitationOperatorsDef2} is in general restricted only to a $a$-dependent subset of the occupied orbitals, depending on the particular symmetry
of the excitation. Excited states  satisfy the normalization condition
\be\lb{ExcitedStateOrthNormDef1}
\sum_p \left(\brket{\phi_p^a}{\phi_p^b} - \brket{\chi_p^b}{\chi_p^a}\right) = \delta_{ab} \;.
\ee
By applying \eqref{ExcitationOperatorsDef1} to the above parametrization
of the excitations, we obtain the following Sternheimer-like equations for the excited states:
\begin{align}\lb{ExcitationOperatorsDef3}
&\left[\Omega_a-(\hnot - \eps_p)\right] \ket{\phi_p^a} = \op Q_0 \op V'[\op E_a] \ket{\psi_p} \;, \nn\\
&\bra{\chi_p^a}\left[-\Omega_a-(\hnot - \eps_p)\right] = \bra{\psi_p} \op V'[\op E_a] \op Q_0  \;.
\end{align}
By writing these equations in a basis set of virtual states we obtain the well-known Casida's eigenvalue equation\cite{CasidaBook}
associated to the excitation energy $\Omega_a$ (see appendix \ref{casida}).
The spectrum is symmetric w.r.t the inversion of the eigenvalues $\Omega_a \rightarrow -\Omega_a$ and, given a specific excitation $\{\phi^a_p,\chi^a_p\}$, 
the associated solution with opposite energy is described by the transposed pair $\{\chi^a_p,\phi^a_p\}$.
We concentrate our analysis on the sector of positive energies, i.e. $\Omega_a > 0$.

@@

The susceptibility functional has therefore a peculiar analytic structure made of multiple threshold energies, one for each of the ionization potentials associated to the occupied states. If an excitation $\op E_a$ has an energy $\Omega_a < |\eps_a|$ with $\eps_a = \mathrm{max}\left(\eps_p\right)_{p\in a}$, it may belong to a discrete spectrum; otherwise it \emph{has} to belong to a continuum of states.
The typical structure of the excitation landscape is depicted in figure \ref{ExcitationLandscape}, once again assuming that the perturbed potential $\op V'[\op E]$ gives rise to a bound state when applied to $\ket{\psi_p}$. Below the first ionization energy $|\eps_h|$ only localized excitations are present. Conversely, in the intermediate energy interval that extends from $|\eps_h|$ to the absolute value of the deepest occupied orbital, other discrete excitations may exist, and are embedded in a continuum of delocalized excitations. For energies higher than $|\eps_1|$ \emph{no localized} excitations are anymore possible,
regardless of the particular behavior of $\op V'$: the excitation landscape in this region contains a continuum set of delocalized states.

\section{Physical relevance of the excitation spectrum. A comparison among localized and continuum sectors}
% \deleted[id=dal]{
% Comments on the features and the physical relevance of the two sector of the excitation spectrum}
% 
% %\subsection{Localized sector}
% 
% %\begin{itemize}
% % \deleted[id=dal]{\item It is composed by a genuine set of discrete elements.}
% % \deleted[id=dal]{\item The values of the excitation energies do not depend on the computational setup (observable features) and so can be compared among different approaches.}
% % \deleted[id=dal]{\item The knowledge of this sector of the excitation spectrum allows us to express only the imaginary part of (proper equation that expresses the response functional written using Sokowsky-Plemely
% % on the susceptibility functional).}
% % \deleted[id=dal]{\item A computational setup in which delocalized degrees of freedom are excluded by design can in principle describe to this sector in a exact way.}
% %\end{itemize}
% 
% \subsection{Continuum sector}
% 
% \begin{itemize}
%    \item Make some statements on the features that a computational setup must have to guarantees that this sector provides converged results. Difference between systematic and localized basis setup. Can we comment about the needs to add ghost atoms for the evaluation of statical polarizability with gaussian basis?
%  % \deleted[id=dal]{\item Not true poles of the susceptibility (but only an effective tool to compute the spectral function).}
%  % \deleted[id=dal]{\item Deeply affected by the computational setup used to represent them. In particular, the presence of a finite computational domain implies a fictitious discretization of the excitation levels and the associated spectrum exhibits an explicit dependence from the size of the domain.}
%  % \deleted[id=dal]{\item This sector of the excitation spectrum is needed to compute imaginary parts above IP and real part for \emph{all} the values of $\omega$. Add examples on the statical polarizability. Can we comment}
% \end{itemize}
% 
% 
% \vspace{1cm}

@@

The analysis of the above section reveals that the excitation spectrum is composed by two distinct sectors.
The first one is realized by the (finite) set of localized excitations with discrete eigenvalues.
Borrowing the terminology introduced in section \ref{SEopenSystem} to the present case, we can affirm that such localized excitations behave as
\emph{observable} quantities and the associated energies possess ideal features of reproducibility:
if the numerical basis employed is complete enough to
represent the states $\{\phi_p^a,\chi_p^a\}$, the energy $\Omega_a$
converges to its reference value.
A computational setup in which delocalized degrees of freedom are excluded by design can, in principle, describe excitations of this sector in a exact way.
